---
title: "HousePrices : Exploratory Data Analysis"
output: 
  html_document:
    toc: true
    toc_depth: 5
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning = FALSE)
```

### Introduction
This document is an analysis of the Kaggle dataset at https://www.kaggle.com/c/house-prices-advanced-regression-techniques

With 79 explanatory variables describing (almost) every aspect of residential homes in Ames, Iowa, this competition challenges data scientists/analysts to predict the final price of each home.

This document is limited to the exploratory analysis of the data, visualizations and attempt to find interesting patterns. The actual machine learning/predictive efforts will be carried out in another document.

```{r LoadLibraries,message=FALSE}
rm(list = ls(all = TRUE))
library(dplyr)
library(ggplot2)
library(gridExtra)
library(tabplot)
library(stringr)
library(VIM)
library(caret)
library(ggmap)
library(Boruta)
options(scipen=999)
```


```{r Data Load,eval=FALSE}
#Will Load files directly from the cleaned and save combi.rds
train <- read.csv("train.csv", stringsAsFactors=FALSE)
test <- read.csv("test.csv", stringsAsFactors=FALSE)
```

```{r CleanData,eval=FALSE}
test$SalePrice=NA
train$IsTrain=TRUE
test$IsTrain=FALSE
combi <- rbind(train,test)
test_Id=test$Id


str(train$Id)


#Analysis of blanks and NAs
#Check NAs
sapply(combi,function(x){sum(ifelse(is.na(x),1,0))}) #This is faster than using colSums below
#colSums(sapply(combi,is.na))
#There are a lot of Variables where NA actually means something. See the data description. Lets analyze them first

#Alley
nrow(combi[is.na(combi$Alley),])
#2721 NAs As per data dictionary NA means no alley/ So lets update NAs to No.
combi$Alley <- ifelse(is.na(combi$Alley),"NO",combi$Alley)

#BsmtQual
nrow(combi[is.na(combi$BsmtQual),])
#81 NA's
combi$BsmtQual <- ifelse(is.na(combi$BsmtQual),"NO",combi$BsmtQual)

#BsmtCond
combi$BsmtCond <- ifelse(is.na(combi$BsmtCond),"NO",combi$BsmtCond)

#BsmtExposure
combi$BsmtExposure <- ifelse(is.na(combi$BsmtExposure),"NO",combi$BsmtExposure)

#BsmtFinType1
combi$BsmtFinType1 <- ifelse(is.na(combi$BsmtFinType1),"NO",combi$BsmtFinType1)

#BsmtFinType2
combi$BsmtFinType2 <- ifelse(is.na(combi$BsmtFinType2),"NO",combi$BsmtFinType2)

#FireplaceQu
combi$FireplaceQu <- ifelse(is.na(combi$FireplaceQu),"NO",combi$FireplaceQu)

#GarageType
combi$GarageType <- ifelse(is.na(combi$GarageType),"NO",combi$GarageType)

#GarageFinish
combi$GarageFinish <- ifelse(is.na(combi$GarageFinish),"NO",combi$GarageFinish)

#GarageQual
combi$GarageQual <- ifelse(is.na(combi$GarageQual),"NO",combi$GarageQual)

#GarageCond
combi$GarageCond <- ifelse(is.na(combi$GarageCond),"NO",combi$GarageCond)

#PoolQC
combi$PoolQC <- ifelse(is.na(combi$PoolQC),"NO",combi$PoolQC)

#Fence
combi$Fence <- ifelse(is.na(combi$Fence),"NO",combi$Fence)

#MiscFeature
combi$MiscFeature <- ifelse(is.na(combi$MiscFeature),"NO",combi$MiscFeature)

#Done. Now check how many NAs left
sapply(combi,function(x){sum(ifelse(is.na(x),1,0))})


combi$Utilities= ordered(combi$Utilities,levels= c("ELO","NoSeWa","NoSewr","AllPub"))
combi$ExterQual= ordered(combi$ExterQual,levels=c("Po","Fa","TA","Gd","Ex"))
combi$ExterCond= ordered(combi$ExterCond,levels=c("Po","Fa","TA","Gd","Ex"))
combi$BsmtQual= ordered(combi$BsmtQual,levels=c("NO","Po","Fa","TA","Gd","Ex"))
combi$BsmtCond= ordered(combi$BsmtCond,levels=c("NO","Po","Fa","TA","Gd","Ex"))
combi$BsmtExposure= ordered(combi$BsmtExposure,levels= c("NO", "No", "Mn","Av","Gd"))
combi$BsmtFinType1= ordered(combi$BsmtFinType1,levels=c("NO", "Unf","LwQ","Rec","BLQ","ALQ","GLQ"))
combi$BsmtFinType2= ordered(combi$BsmtFinType2,levels=c("NO", "Unf","LwQ","Rec","BLQ","ALQ","GLQ"))
combi$HeatingQC= ordered(combi$HeatingQC,levels=c("Po","Fa","TA","Gd","Ex"))
combi$KitchenQual= ordered(combi$KitchenQual,levels=c("Po","Fa","TA","Gd","Ex"))
combi$FireplaceQu= ordered(combi$FireplaceQu,levels=c("Po","Fa","TA","Gd","Ex","NO"))
combi$GarageFinish= ordered(combi$GarageFinish,levels=c("NO", "Unf","RFn","Fin"))
combi$GarageQual= ordered(combi$GarageQual,levels=c("NO","Po","Fa","TA","Gd","Ex"))
combi$GarageCond= ordered(combi$GarageCond,levels=c("NO","Po","Fa","TA","Gd","Ex"))
combi$PoolQC= ordered(combi$PoolQC,levels=c("NO","Fa","TA","Gd","Ex"))
combi$Fence= ordered(combi$Fence,levels=c("NO","MnWw","GdWo","MnPrv","GdPrv"))

str(combi)
summary(combi)


#FOr chars I will analyze if there is some sort of ordering. If yes then will convert them
#to ordered factor, else they will be converted directly to factors. I will check the data dictionary
#and will choose those chars which can be ordered.


combi$LotShape <- factor(combi$LotShape,levels=c("IR1","IR2","IR3","Reg"),ordered=TRUE)
combi$CentralAir <- ordered(combi$CentralAir,levels=c("N","Y"))
combi$PavedDrive <- ordered(combi$PavedDrive, levels=c("N","P","Y"))

#Convert all other chars to factors
combi <- combi %>%
  mutate_if(is.character,as.factor)

#There are also some variables which are marked as int but are actually categorical. So we will
#Convert those also to factors
combi$MSSubClass <- as.factor(combi$MSSubClass)

#Age of property
#Age of property will be a better indicator than year built and year sold #Assuming month of year built is 1
combi$AgeBuilt <- (combi$YrSold-combi$YearBuilt)*12 +combi$MoSold


#Change other years into age
combi$AgeRemodAdd <- (combi$YrSold-combi$YearRemodAdd)*12
combi$AgeGarageBlt <- (combi$YrSold-combi$GarageYrBlt)*12

#Drop the YrSold,YearBuilt,"YearRemodAdd","GarageYrBlt" and MoSold variables
#combi[,c("YrSold","YearBuilt","MoSold","YearRemodAdd","GarageYrBlt")] <- NULL
#Drop ID
combi$Id <- NULL


#KNN Imputation
# Right now, I will do a quick kNN imputation to set a benchmark
combi_kNN <- select(combi,-c(SalePrice,IsTrain))
#combi_kNN <- combi[,!colnames(combi) %in% c("SalePrice","IsTrain")] #Removed these three columns

combi_knn_imputed <- kNN(data=combi_kNN,k=54) #Not giving variable names,so it will impute NAs in all vars. k=sqrt(2919)

sapply(combi_knn_imputed,function(x){sum(ifelse(is.na(x),1,0))})
#All NAs have been removed

str(combi_knn_imputed)
#Additional variables have been created. We can remove those


combi_knn_imputed <- subset(combi_knn_imputed,select = MSSubClass:AgeGarageBlt )
str(combi_knn_imputed)

combi <- data.frame(combi_knn_imputed,combi[,c("SalePrice","IsTrain")]) 

#Now check for any blanks
sapply(combi,function(x){sum(ifelse(str_trim(x)=="",1,0))}) #Check no. of blanks in the features
#No blanks. Everything fine.

tableplot(combi[combi$IsTrain==TRUE,],sortCol = SalePrice,nBins = 10,select=c(SalePrice,2:5),nCols = 2)

summary(combi$LotArea)
summary(combi)

#Save clean data
saveRDS(combi, "combi.rds")
train_cleaned <- write.csv(x=combi[combi$IsTrain==TRUE,],file = "train_cleaned.csv",row.names = FALSE)
test_cleaned <- write.csv(x=combi[combi$IsTrain==FALSE,],file = "test_cleaned.csv",row.names = FALSE)
```


Load Data from cleaned CSVs
```{r LoadCleanedData}
combi <- readRDS("combi.rds")
```

Now for some EDA
```{r}
#Now for some analysis and vizualization
train_data <- combi[combi$IsTrain==TRUE,]
test_data <- combi[combi$IsTrain==FALSE,]
```

Checking the characterstic of Sale Price
```{r EDA}
summary(train_data$SalePrice)
ggplot(train_data,aes(SalePrice)) +
  geom_density(col="orange")
boxplot.stats(train_data$SalePrice)
```

About 61 SalesPrices are outlier.Excluding the outliers, the distribution of SalesPrice is roughly Normal.

```{r CompleteTablePlot}
tableplot(dat = train_data,sortCol = SalePrice,nCols = 7)
```

The plots above show the relationship between Sale Price and the other features. Lets quickly check out some important relationships

1.  There does not seem to be a relationship between Sale Price and Lot Area or Lot Frontage
2.  There seems to be a strong positive corelation between Sale Price and Overall Quality
3.  Newly built/remodelled htouses have higher Sale Prices.
4.  It pays to have a Good/EXcellent Exterior and basement Quality.
5.  Poured concrete foundation houses have higher prices.
6.  Houses having larger basements command higher sales price.
7.  Not surprisingly, actual area of the house (ground floor area, 1st floor area) is related to the Sales Price. As noted above, the Lot area is not a strong indicator of the price.
8.  Bigger garages command higher prices.
9.  Also interesting to note is that Prices of houses being sold under abnormal conditions(such as foreclosures) are typically lower.


   
No. of houses sold as a function of Year Built  
  
```{r}
ggplot(train_data,aes(x=YearBuilt)) +
  geom_bar(fill="orange") +
  labs(x="Year",
       y="No. of houses sold",
       title="No. of Houses Sold vs Year Built")
```

   
Avearge Sale Price over the years
  
```{r}
ggplot(train_data) +
  geom_histogram(aes(x=YrSold,y=SalePrice),stat="summary",fun.y=mean,fill="tomato1",labels=TRUE)
```


The Average Sales price have more or less remain stagnant over the past few years

  
  
Some key characterstict of High Sale Price houses.
  
Neighbourhoods with higher Sale Prices
```{r HighPrices}

neighborhood_df <- train_data[order(desc(train_data$SalePrice)),]
#neighborhood_df$Neighborhood <- fct_inorder(neighborhood_df$Neighborhood,ordered = TRUE)
ggplot(neighborhood_df) +
  geom_bar(aes(x=Neighborhood,y=SalePrice,fill=MSZoning),stat="summary",fun.y=median) + 
  geom_hline(aes(yintercept=median(SalePrice)),col="red") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x="Neighborhood",
       y="Median Sales Price",
       title="Sales Price across Neighborhoods")
  
rm(neighborhood_df)
```
  
1.  High cost houses are typically from these neighbourhoods:
    a)  Crawford
    b)  Northridge Heights
    c)  Somerset
The red line is the median Sales Price

```{r AverageGrowthRate}
avggrowth_df <- train_data %>%
                group_by(Neighborhood,YrSold) %>%
                select(Neighborhood,YrSold,SalePrice) %>%
                summarize(avg_annual_price=mean(SalePrice))

ggplot(avggrowth_df,aes(x=YrSold,y=avg_annual_price)) +
         geom_line(aes(col=Neighborhood))

high_low_growth_df <- avggrowth_df %>%
                   group_by(Neighborhood) %>%
                    summarize(avg_price=mean(avg_annual_price)) %>%
                    select(Neighborhood,avg_price) %>%
                    arrange(desc(avg_price)) 

high_low_growth_df <- rbind(head(high_low_growth_df,5),tail(high_low_growth_df,5))

ggplot(avggrowth_df[avggrowth_df$Neighborhood %in% high_low_growth_df$Neighborhood,]) + 
      geom_line(aes(x=YrSold,y=avg_annual_price,col=Neighborhood)) +
  labs(x="Year of Sale",
       y="Avearge Sale Price",
       title= "Average Sale Price of five highest and lowest priced neighborhoods")
rm(avggrowth_df)    
rm(high_low_growth_df)
```


```{r}
table(train_data$SaleCondition)
ggplot(train_data) +
  geom_bar(aes(x=YrSold,fill=SaleCondition),position="fill") +
  labs(x="Year Sold",
      y="Proportion",
      title="Sale Condition as proportion of total Sales")

t.test(train_data[train_data$YrSold==2007 & train_data$SaleCondition=="Abnorml","SalePrice"],train_data[train_data$YrSold==2008 & train_data$SaleCondition=="Abnorml","SalePrice"])
```

The proportion of Normal sale as a fraction of total sale has been increasing. I dont think there is data for enough years to conclude from this chart whether the proportion of Abnormal sales increased due to the Financial crisis. The t test for the avearge Sale Price of "Abnormal" sales in 2007, 2008 indicate that there is no price difference between the two years.


Boruta Feature Importance

```{r Boruta,eval=F}
Sys.time()
  boruta_features <-  Boruta(data=train_data,SalePrice~.,doTrace=2)
Sys.time()
boruta_features$finalDecision

```

